---
title: "Sierra Amphibians Visualization"
description: |
  This report creates a final visualization (two-part graph) of frog count data for the mountain yellow-legged frog using Sierra Lakes Inventory Project data.
author:
  - name: Karina Johnston
    url: {}
date: 2022-03-13
output:
  distill::distill_article:
    self_contained: false
    toc: yes
    code_folding: hide
---


```{r setup, include = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Attach packages
library(here)
library(tidyverse)
library(readxl) # to read in excel files
library(lubridate) # to update dates
library(cowplot) # to merge the graphs

```

## Introduction

The Sierra Lakes Inventory Project (SLIP) was a research endeavor that ran from 1995-2002 and has supported research and management of Sierra Nevada aquatic ecosystems and their terrestrial interfaces. They described the physical characteristics of and surveyed aquatic communities for > 8,000 lentic water bodies in the southern Sierra Nevada, including lakes, ponds, marshes, and meadows (Knapp et al. 2020). The objective of this report was to create a final visualization (two-part graph) of the frog count data for the mountain yellow-legged frog (*Rana muscosa*). 


```{r}
# Read in the data
amphibians <- read_excel(here("data", "sierra_amphibians.xlsx")) 

```


## Data Wrangling

Filter the data to be only the mountain yellow-legged frog (*Rana muscosa*, code RAMU) and keeping only the life stages of adult, subadult, and tadpole. Mutate the columns to add a new one with "year" using the `lubridate` package. 

```{r}
# Subset the data

amph_sub <- amphibians %>% 
  filter(amphibian_species == "RAMU") %>% # choose only Rana muscosa
  filter(amphibian_life_stage != "EggMass") %>% # drop the Eggmass category

  # updates the date column to be Date format and adds a year column 
  mutate(date = ymd(survey_date)) %>% 
  mutate(year = year(survey_date)) 

```


```{r}
# Summarize the counts by year

amph_summary <- amph_sub %>% 
  group_by(year, amphibian_life_stage) %>% 
  summarize(total = sum(amphibian_number))

```

A second set of data wrangling to count only the adult and subadult mountain yellow-legged frogs and to rename the lake column in the dataframe.

```{r}
# Subset the data

amph_adult_sub <- amph_sub %>% 
  filter(amphibian_life_stage != "Tadpole") %>% # delete Tadpoles
  mutate(lake = "Lake") %>% # add a new column "Lake"
  unite("lake_id_new", lake, lake_id, sep = " ", remove = FALSE)
# use unite to merge the columns and create a new one to sort by

# Summarize the counts by lake id 
amph_adult_summary <- amph_adult_sub %>% 
  group_by(lake_id_new) %>% 
  summarize(total = sum(amphibian_number)) 

```


## Data Visualization

Create (A) a graph of total mountain yellow-legged frog (*Rana muscosa*) count each year across all water bodies, by life stage excluding the ‘EggMass’ level, and (B) a column graph containing total counts (over all years of the study) of combined adult and subadult endangered mountain yellow-legged frogs observed in the five lakes with the greatest total observed counts.

```{r}
# Creates a new ggplot figure visualizing amph_summary

ramu_count <- ggplot(data = amph_summary, aes(x = year, y = total)) +
  geom_col(aes(fill = amphibian_life_stage), 
           color = "black") +

  # changes fill colors to be more aesthetic
  scale_x_continuous(n.breaks = 5) + # to clean up the x-axis year numbering
  scale_fill_manual(values = c("blue", "light blue", "yellow"), 
                    name = "Life Stage") +
  scale_y_continuous(labels = scales::comma, n.breaks = 8) +

  # updates the axis labels
  labs(x = "Year", y = "Total Number of Frogs") +

  theme_classic() # clean theme

```


```{r, fig.width = 9, fig.height = 4, fig.cap = "**Figure 1.** Sierra Lakes mountain yellow-legged frog (*Rana muscosa*) counts: (A) by year and life stage across all waterbodies, and (B) combined adult and subadult counts for the top five waterbodies (Data: Knapp et al. 2020)."}

# Create a second visualization bar graph with totals by lake

ramu_adult <- amph_adult_summary %>% 
  
  # to reorder and cut the top 5 lakes from the dataframe
  mutate(lake_id_new = fct_reorder(lake_id_new, desc(total))) %>% 
  slice_max(order_by = total, n = 5) %>% 
  
# Create the new ggplot using the adult frog data
 ggplot(aes(x = lake_id_new, 
                         y = total)) + 
  geom_col(fill = "blue", 
           color = "black",
           width = 0.5) + 
  theme_classic() + 
#  coord_flip() +
  labs(x = "Lake ID", y = "Total Number of Frogs") +
  scale_y_continuous(labels = scales::comma, n.breaks = 10) 


# Use the cowplot package to merge the plots into one figure
plot_grid(ramu_count, ramu_adult, 
          labels = c("A", "B"), 
          label_size = 12)

```


## Citations

Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3

R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL   https://www.R-project.org/.

Wickham, H. et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

Wickham, H. and J. Bryan (2019). readxl: Read Excel Files. R package version 1.3.1. https://CRAN.R-project.org/package=readxl

