---
title: "Sugar Maples and Soil Acidification"
description: |
  This report explores the characteristics of sugar maple seedlings in a Calcium (Ca)- treated watershed at Hubbard Brook Experimental Forest as compared  to the characteristics of seedlings growing on reference sites.
author:
  - name: Karina Johnston
    url: {}
date: 2022-02-20
output:
  distill::distill_article:
    self_contained: false
    toc: yes
    code_folding: hide
---


```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Attach packages
library(tidyverse)
library(here)
library(kableExtra)
library(effsize)

```


```{r}
# Read in the data
maple_data <- read_csv(here("data", "maples.csv"))
```

### Introduction

This report explores the characteristics of sugar maple seedlings in a Calcium (Ca)- treated watershed at Hubbard Brook Experimental Forest as compared  to the characteristics of seedlings growing on reference sites (Juice and Fahey 2019). The reference location has been impacted by acid rain, which leaches calcium, an important plant nutrient, while the other watershed has had calcium additions. 

Preliminary summary statistics are calculated for both survey years (2003-2004) and subsequent analyses are conducted on the 2003 year only. The primary question analyzed is: "Is there a significant difference in stem dry masses for sugar maples, recorded in 2003, between the reference and calcium-treated watersheds?"

### Summary Data Table - 2003-2004 

Table 1 contains a summary of of stem mass (g) and stem length (mm) data for Hubbard Brooks LTER sugar maples in a calcium-treated watershed (W1) and a reference watershed (SD = 1 standard deviation) (Data: Juice and Fahey 2019).


```{r}
# this code chunk creates a new dataframe with summary statistics to output a final table

maple_table <- maple_data %>% # create a new dataframe
  group_by(watershed, year) %>% # group by variables

  # output summary statistics into the dataframe table
  summarize(mean_stem_mass = round(mean(stem_dry_mass, na.rm = TRUE), 3),
            sd_stem_mass = round(sd(stem_dry_mass, na.rm = TRUE), 3),
            mean_stem_length = round(mean(stem_length, na.rm = TRUE), 2),
            sd_stem_length = round(sd(stem_length, na.rm = TRUE), 2),
            sample_size = n()
  )

# these functions update the table to be cleaner, easier to read, and visually appealing / finalized
kable(maple_table,
      col.names = c("Watershed", # changes the column names
                    "Year", 
                    "Mean Stem Mass (g)",
                    "SD Stem Mass (g)",
                    "Mean Stem Length (mm)",
                    "SD Stem Length (mm)",
                    "Sample Size"),
      align = "lcccccc", # changes the alignment of the columns
    
# adds a caption
      caption = "Table 1. Summary of stem mass (g) and stem length (mm) data for Hubbard Brooks LTER sugar maples in a calcium-treated watershed (W1) and a reference watershed (SD = 1 standard deviation) (Data: Juice and Fahey 2019).") %>% 

# better lines within the table
kable_classic(full_width = FALSE)
```


### Data Analyses - 2003 Only

Figure 1 visualizes both the raw data and summary statistics for sugar maple stem dry masses (g) compared between a reference (acid rain) watershed and a calcium-treated watershed (Data: Juice and Fahey 2019).

```{r}
# create a new subset dataframe using sugar maple data from only 2003

maple_sub <- maple_data %>% 
  filter(year == "2003")

```

<br>

```{r, fig.cap = "**Figure 1.** Sugar maple stem dry masses (g) compared between a reference (acid rain) watershed and a calicium-treated watershed. The graph contains both raw data and summary boxplot statistics (Data: Juice and Fahey 2019).", fig.align = "center"}

# create a final visualization comparing stem dry mass by watershed

# to change label names and colors
xlabs <- c("Reference", "Calcium Treated")
boxcolor <- c("cyan4", "orange")


# graph showing stem dry mass by watershed
ggplot(data = maple_sub, aes(x = watershed, y = stem_dry_mass)) +
  geom_boxplot(aes(fill = watershed)) +
  geom_jitter(data = maple_sub, # to also visualize/include the raw data
              aes(x = watershed, y = stem_dry_mass), 
              color = "black", # points will be black
              alpha = 0.5, # increase transparency
              position = position_jitter(width = 0.15)) + # to narrow the width of the jitter
  theme_classic(base_size = 12) +
  labs(x = "Watershed", y = "Stem Dry Mass (g)") +

  scale_x_discrete(labels = xlabs) + # make labels more readable than a code
  scale_fill_manual(values = boxcolor) + # change the fill color
  scale_y_continuous(breaks = waiver(), # increase the number of y-axis breaks
                     n.breaks = 8) +
  guides(fill = FALSE) # removes unnecessary legends

```

<br>

#### Exploratory Visualizations

This section creates several exploratory (unfinalized) visualizations to determine if a parametric statistical test is appropriate. 

```{r}
# exploratory visualizations

# exploratory histogram
ggplot(data = maple_sub, aes(x = stem_dry_mass)) +
  geom_histogram(bins = 16, aes(fill = watershed), color = "black") +
  facet_wrap(~ watershed) +
  labs(x = "Stem Dry Mass (g)", y = "Count") +
  theme_classic()

# exploratory quantile-quantile plot 
ggplot(data = maple_sub, aes(sample = stem_dry_mass)) +
  geom_qq(aes(color = watershed)) +
  facet_wrap(~ watershed) +
  theme_classic()

```

The histogram of stem dry mass (g) shows that the data are relatively normally distributed, and the quantile-quantile plot shows a relatively linear comparison of the sample to theoretical distributions. Because the data appear normal, a Welch's t-test for comparison of two means is an appropriate statistical analysis. 

```{r}
# Conduct a t-test as the statistical analysis

mass_reference <- maple_sub %>% # creates a new dataframe
  filter(watershed == "Reference") %>% # to filter reference records
  drop_na(stem_dry_mass) %>% # to drop the NAs
  pull(stem_dry_mass) # to pull a vector of just mass (easier for t-test)

mass_treated <- maple_sub %>% # creates a new dataframe 
  filter(watershed == "W1") %>% # to filter treated records
  drop_na(stem_dry_mass) %>% # to drop the NAs
  pull(stem_dry_mass) # to pull a vector of just mass (easier for t-test)

# run a t-test to determine significant difference
mass_ttest <- t.test(mass_reference, mass_treated)
# mass_ttest # output the t-test results

# run the cohen d test (for effect size)
mass_cohen_d <- cohen.d(mass_reference, mass_treated)
# mass_cohen_d # output the cohen d results

```

Mean stem mass (g) for the reference watershed in 2003 (`r maple_table$mean_stem_mass[1]` $\pm$ `r maple_table$sd_stem_mass[1]`) was found to be significantly lower than mean stem mass (g) from the calcium treated watershed (`r maple_table$mean_stem_mass[3]` $\pm$ `r maple_table$sd_stem_mass[3]`) (mean $\pm$ 1 standard deviation) (Welch t-test: t(`r round(mass_ttest$parameter, 1)`) = `r round(mass_ttest$statistic, 3)`; *p* = `r ifelse(mass_ttest$p.value < 0.001, paste("< 0.001"), paste("=", mass_ttest$p.value))`). Cohen's *d* was also used to assess effect size which incorporates the sample size for each watershed (Cohen's *d* = `r round(mass_cohen_d$estimate, 2)`). Since the effect size was large, it is likely that a practical difference exists between the stem mass of the two watersheds, with the stems from the calcium treated watershed larger on average. 

In this experiment, the addition of calcium in the treated watershed appeared on average to improve sugar maple seedling growth as compared to the acid-rain impacted (reference) watershed. While this does not *prove* that calcium additions increased the sugar maple growth patterns, the results indicated substantially larger stem mass with the calcium additions in this experiment. 


### Citations

Juice, S. and T. Fahey. 2019. Health and mycorrhizal colonization response of sugar maple (Acer saccharum) seedlings to calcium addition in Watershed 1 at the Hubbard Brook Experimental Forest ver 3. Environmental Data Initiative. https://doi.org/10.6073/pasta/0ade53ede9a916a36962799b2407097e 

R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Torchiano M (2020). effsize: Efficient Effect Size Computation. doi:
10.5281/zenodo.1480624 (URL: https://doi.org/10.5281/zenodo.1480624), R package
version 0.8.1, <URL: https://CRAN.R-project.org/package=effsize>.

Wickham, H. et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

Zhu, H. (2021). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax. R package version 1.3.4. https://CRAN.R-project.org/package=kableExtra


